Embracing the Laws of Physics: \\ Reversible Models of Computation

### 2017-12-22 survey/survey.tex
More in Data II. An explanation that satisfied me for Pi seems to take more space than expected.  Ah well, it is coming along.

@@

Nevertheless, the Curry-Howard correspondance still has some
force. We know that the inhabitants of types formed with
with $\bot, \top, \uplus, \times$ are a commutative semiring. What we
want to know is, which types are equivalent? From a commutative semiring
perspective, this amounts to asking what terms are equal?
We have a set of generators for those equations, namely those
in~(\ref{eq:csemiring}). What we thus need is to create $8$ pairs
of mutually inverse functions which witness these identities.
For concreteness, we show the signatures in Figure~\ref{pi-terms}.

>>>>

\begin{figure}[t]
\[
\begin{array}{rrcll}
\idc :& \tau & \simeq & \tau &: \idc \\
\\
\identlp :&  \bot \uplus \tau & \simeq & \tau &: \identrp \\
\swapp :&  \tau_1 \uplus \tau_2 & \simeq & \tau_2 \uplus \tau_1 &: \swapp \\
\assoclp :&  \tau_1 \uplus (\tau_2 \uplus \tau_3) & \simeq & (\tau_1 \uplus \tau_2) \uplus \tau_3 &: \assocrp \\
\\
\identlt :&  \top \times \tau & \simeq & \tau &: \identrt \\
\swapt :&  \tau_1 \times \tau_2 & \simeq & \tau_2 \times \tau_1 &: \swapt \\
\assoclt :&  \tau_1 \times (\tau_2 \times \tau_3) & \simeq & (\tau_1 \times \tau_2) \times \tau_3 &: \assocrt \\
\\
\distz :&~ \bot \times \tau & \simeq & \bot ~ &: \factorzl \\
\dist :&~ (\tau_1 \uplus \tau_2) \times \tau_3 & \simeq & (\tau_1 \times \tau_3) \uplus (\tau_2 \times \tau_3)~ &: \factor
\end{array}
\]
\caption{Type isomorphisms.}
\label{type-isos}
\end{figure}

Nevertheless, the Curry-Howard correspondance still has some
force. We know that the inhabitants of types formed with
with $\bot, \top, \uplus, \times$ are a commutative semiring. What we
want to know is, which types are equivalent? From a commutative semiring
perspective, this amounts to asking what terms are equal?
We have a set of generators for those equations, namely those
in~(\ref{eq:csemiring}). What we thus need is to create $8$ pairs
of mutually inverse functions which witness these identities.
For concreteness, we show the signatures in Figure~\ref{type-isos}.
As these all come in symmetric pairs (some of which are self-symmetric),
we give names for both directions. The names use $\iso$, $+$ and $*$
as these will be what we use when we reflect these isomorphism into
a programming language, which we proceed to do next.

@@

We now have in our hands our desired denotational semantics: we want
to create a language $\Pi$ such that the types and type combinators
map to $\bot, \top, \uplus, \times$, and such that we have ground
terms whose denotation are all $16$ type isormorphisms of
Figure~\ref{type-isos}. This is completely straightforward, as
we can simply do this literally. To make the analogy with 
commutative semirings stand out even more, we will use $0, 1, +$, and $*$
at the type level, and will denote ``equivalence'' by $\iso$.
Thus Figure~\ref{pi-terms} shows the ``constants'' of the language.

Of course, one does not get a programming language with just (typed)
constants!
